---
title: "Exercise 4"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---



```{r}
library(Banksy)
library(igraph)
library(ggspavis)
library(BayesSpace)
library(nnSVG)
library(HDF5Array) # For loading object
library(patchwork)
library(SpatialExperiment)
library(scran) 
library(scater)
```

# Dimensionality Reduction

To avoid large computational time, we will load the new subset containing 2000 HVG

```{r}
spe.hvg <- loadHDF5SummarizedExperiment(dir="results/day1", prefix="01.3_spe_tmp_tmp")
spe <- loadHDF5SummarizedExperiment(dir="results/day1", prefix="01.3_spe_tmp_tmp")
```


### Non-spatially aware:

#### PCA:

```{r}
spe <- runPCA(spe, ncomponents = 20, subset_row = rowData(spe)$hvg)

# Visualize the explained variance by each principal component.
# This helps in determining how many PCs capture most of the variance.
plot(attr(reducedDim(spe, "PCA"), "percentVar"),
  xlab = "PC", ylab = "Proportion of variance explained",
  main = "PCA Scree Plot"
)


spe <- runPCA(spe, ncomponents = 10, subset_row = rowData(spe)$hvg)

## Visualize PCs on the tissue slice coordinates
pcs <- reducedDim(spe, "PCA")
#pcs <- pcs[, seq_len(10)]
ps <- lapply(colnames(pcs), \(.) {
    spe[[.]] <- pcs[, .]
    plotCoords(spe, annotate = .)
}) 
patchwork::wrap_plots(ps, nrow = 2) & 
  theme(legend.key.width = unit(0.4, "lines"), 
        legend.key.height = unit(0.8, "lines")) & 
  scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")))

```

### Spatially aware

#### Banksy

```{r}
# Clustering with Banksy
set.seed(112358)

# 'Banksy' parameter settings
k <- 20   # consider first order neighbors
l <- 0.8 # use little spatial information
a <- "logcounts"
xy <- c("array_row", "array_col")
# restrict to selected HVG features
dec <- modelGeneVarByPoisson(spe)
hvg <- getTopHVGs(dec, n = 2000) # Selecting top 2000 HVGs
tmp <- spe[hvg, ]
# compute spatially aware 'Banksy' PCs
tmp <- computeBanksy(tmp, assay_name=a, coord_names=xy, k_geom=k)
tmp <- runBanksyPCA(tmp, lambda=l, npcs=10)
reducedDim(spe, "PCA_banksy") <- reducedDim(tmp)

## Visualize PCs on the tissue slice coordinates
pcs <- reducedDim(spe, "PCA_banksy")
#pcs <- pcs[, seq_len(10)]
ps <- lapply(colnames(pcs), \(.) {
    spe[[.]] <- pcs[, .]
    plotCoords(spe, annotate = .)
}) 
patchwork::wrap_plots(ps, nrow = 2) & 
  theme(legend.key.width = unit(0.4, "lines"), 
        legend.key.height = unit(0.8, "lines")) & 
  scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")))
```

### Run UMAP

run UMAP (for visualization purposes only) based on both, spatially aware and not spatially aware dimensionality reduction

```{r}
# UMAP
spe <- runUMAP(spe, dimred="PCA", name="UMAP_tx")
spe <- runUMAP(spe, dimred="PCA_banksy", name="UMAP_banksy")
```

```{r}
# Visualise
plotReducedDim(spe, dimred = "UMAP_tx", colour_by = "sum") 
plotReducedDim(spe, dimred = "UMAP_banksy", colour_by = "sum")
```



# Clustering

### Non-spatial 

```{r}
# build cellular shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(spe, use.dimred="PCA", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.6)
table(spe$Leiden <- factor(k$membership))
```


### Spatial 

#### Probabilistic: BayesSpace

```{r}
# prepare data for 'BayesSpace'
# skipping PCA (already computed)
spe <- spatialPreprocess(spe, skip.PCA=TRUE)
# perform spatial clustering with 'BayesSpace'
# using 'd=10' PCs and targeting 'q=8' clusters
spe <- spatialCluster(spe, q=8, d=10, nrep=1e3, burn.in=100) 
table(spe$BayesSpace <- factor(spe$spatial.cluster))

```


### Banksy

We have computed a principal components using Banksy mathod already in the previous step. We will now used them (use Banksy dimentionality reduction space) to perdoem SNN graph-based Leiden clustering based on them, in order to obtain spatially-aware clusters.
When building the SNN graph, we will use the same parameters we have used before for the non-spatially aware graph, but using banksy PCs.

```{r}
# perform SNN graph-based clustering on 'Banksy' PCs using
g <- buildSNNGraph(spe, use.dimred="PCA_banksy", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=0.6)
table(spe$Banksy <- factor(k$membership))

```

```{r}
plotVisium(spe, zoom=T,  annotate = "Banksy")
```


# Visualisation of the different clustering methods:

```{r}
ks <- c("BayesSpace", "Leiden", "Banksy")#, "CellCharter")
plots <- c(lapply(ks, \(.) {
    plt <- plotVisium(spe, annotate=., zoom = T)
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt
}) ,  plotVisium(spe, zoom = T,  spots=FALSE ))
plots |>  wrap_plots(nrow=2) &
    #scale_color_manual(values=unname(pals::trubetskoy())) &
    theme(legend.key.size=unit(0, "lines"), legend.justification="left")
```

```{r}
# plot selected clusters in order of frequency,
# highlighting cells assigned to cluster 'k'
lapply(tail(names(sort(table(spe$Banksy))), 8), \(k) {
    spe$foo <- spe$Banksy == k
    spe <- spe[, order(spe$foo)]
    plt <- plotCoords(spe, annotate="foo")
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt + ggtitle(k)
}) |>
    wrap_plots(nrow=3) &
    scale_color_manual(values=c("lavender", "purple")) &
    theme(plot.title=element_text(hjust=0.5), legend.position="none")
```


